Human Factors Influencing Visual Statistical Inference

### 2013-06-09 sociology_paper.tex
discussion on location effect updated

Figure~\ref{fig:tower} displays a lineup of 20 plots where one of the plots is observed data, while the remaining 19 plots are rendered from data generated under a null model. Which one of the 20 plots is the most different from the others? When asked this question, 12 of 72 observers picked the same plot (with number equal to the result of $3^2+4$), with reasons given being 'asymmetry' (36\%), 'trend' (26\%) or `outliers' (16\%).  The corresponding $p$-value is 0.00023, indicating sufficient evidence to reject the null hypothesis. 
, indicating that we have enough evidence to reject the null hypothesis.
What does this mean, though? For that, we need to know the context of the data and we need to have more information about the generation of the null plots. For that, we need to know the context of the data and we need to have more information about the generation of the null plots. This example investigates the results from the 2012 US presidential election in comparison to the poll results just prior to the election. 
(Although this example is more simplistic than most of the tests conducted to date, it will serve the purpose of illustrating the lineup protocol.) The data is looking at the difference in poll results between the two (major) presidential candidates, Obama and Romney, for all states. Each panel in Figure~\ref{fig:elect-1} shows an `electoral building' \citet{nytimes} where each state in the union is represented by a rectangle. This difference is plotted horizontally, and the height of each box corresponds to the state's electoral votes. Color indicates party affiliation.  The null hypothesis is ``that the election results were consistent with the polls''. The polling results provides the null model from which data is simulated. Because each poll has a margin of error, this is used to simulate different scenarios that might have resulted on election day, if the polls were on target. A null data set is then generate as a set of draws from normal distribution, with mean equal to the difference in poll percentage of the latest state poll results, and standard deviation equal to 2.5, approximating a margin of error of 5\%. These samples are plotted as electoral buildings, and the plot made with the results from the election is placed randomly among them in a lineup of size $5\times 4$. If the null hypothesis is true then the actual data plot should look just like any of the other plots, and not be identified by an observer.
 \hh{Figure~\ref{fig:tower} shows a plot of the electoral building with added context information and labels.}
 %A party needs 270 electoral votes or more to win the presidential election. 
%\begin{figure}[htbp] 
%   \centering
%   \includegraphics[width=0.8\linewidth]{electoral-5-13.pdf} 
%   \caption{Which plot looks the most different from the other plots?  Lineup \#1 - electoral building; another set of four lineups can be found in figs.~{fig:elect.2} through~{fig:elect.5}}
%   \label{fig:elect.1}
%\end{figure}

@@

\subsection{Location of Actual Plot in the Lineup} The actual data plot is placed in a random spot in a lineup. While evaluating the lineup some people may start looking from some specific part of the lineup. With the help of eye-tracking equipment \citet{zhao:2012} track the observers' eyes to see how they go through the plots in a lineup to come to their answers. The results suggest that people have particular methods of reading lineups. Some people read lineups from left to write direction while some read from upward to downward. Some people start looking from the center of the lineup while others start from the top left corner. In the earlier phase of the exercise, the observer tend to scan the plots and then start comparing plots to make a final decision. Beside right-left or up-down directions observer show some diagonal movement too.

Given a specific pattern of eye movement of the observer in examining the lineup, the location of actual plot in the lineup may have some effect. Those who start exploring from the left top corner may get first glance of the actual plot if it is on that location. This should give the observer more time to compare it with the rest of the null plots. Those who start looking from the center may then scan towards right or left direction and thus don't have the chance to scan the null plots continuously as they could if they would start from a corner. Thus they should scan the center plot over and over again while scanning the left or right side plots. If the data plot is in the center location it may have multiple chance to be examined and hence be identified.

@@

\paragraph{Design for Location effect:} The experiment was set up based on the gene expression data with two groups. The groups are defined based on the presence or absence of iron sufficiency. The null hypothesis is that there is no difference between this two groups. Null plots were obtained by randomizing the group structure of the data. The actual plot was randomly placed in five different locations. For each location, five different sets of null plots were used to produce five different lineups. This produced 25 different lineups. Similar procedure was applied to obtain 25 lineups to test the interaction between Genotype and Empty Vector (EV).

@@

Table \ref{tbl:model_result_time} presents the results of Model \eqref{eqn:trend_time}. The parameter $\alpha$ for fixed effect covariate attempt is highly significant in all the experiments. The negative estimates suggest that later attempt took less time for an evaluation. This suggest that even though observers did not improve the performance over attempt, they became efficient in responding faster. The parameter $\alpha_1$ for first attempt is also highly significant. The positive estimates of $\alpha_1$ indicates that first attempt made by an observer required much more times than other attempts. It is because for initial attempt the observer might have gone through instruction and became familiar with the experimental environment.

@@

@@

A plot of residuals against the attempts $A_k$ should reveal the trend if there is any. Figure \ref{fig:learning_trend} shows such plots for three experiments naming experiment 5, 6 and 7. For all the experiments we see some increasing trend of the residuals. This could be attributed to the exhaustion of subjects' attempt after some earlier evaluations are made.  But none of apparent trend are statistically significant.

Learning trend is not only related to the proportion correct responses. It may be possible that people learn to answer fast. We examine this by the time taken for each evaluation. We fit a mixed effect linear model with time taken considering lineup and subjects as random effect. The resulting residuals are plotted in Figure \ref{fig:learning_trend_time}. It appears that in the later attempts observers take less time to evaluate a lineup. Even though it does not tell much about improving their performance over attempts, but it does indicate the improvement of their evaluation skill. Adding this information with what we observe in Figure \ref{fig:learning_trend}, we can say that the observers' improved skills could not contribute to their performance. It only help them finish the task faster in later attempts.

@@

We set up a Turk experiment to examine whether there is any difference of performance based on the location of actual data plot in the lineup. For this five locations of a lineup were randomly chosen to put the same actual data plot. The locations are 2,9,12,16,20 for Interaction effect and 1,8,12,17,20 for Genotype effect. For a lineup with specific data plot location, five sets of null plots are used giving 5 lineups for each location. In total we have 25 lineups for Interaction effect and 25 lineups for Genotype effect. These 50 lineups are then evaluated by 100 people recruited from Amazon Mechanical Turk. Each person evaluated 3 lineups one from Interaction, one from Genotype and the third one was a test lineup. The test lineup is used to process the payment and examine the quality of the data and the responses of the test plot is not added to our analysis.

The proportion of correct responses by the turk observers are shown in Figure \ref{fig:location_effect}. Notice that the data plot is same for each location but we see some variability of performance based on different null plots. This may happen if some null plots appear to be more similar to the actual plot while the others are not making some lineups harder than others even though the actual data plot is same. This pattern is evident in the Figure \ref{fig:location_effect} as we see proportion correct for null plot 5 is consistently above the null plot 1 for each of the locations. But this variability could be controlled at some extent by collecting more data for each location. Since the variability is larger when we have small number of responses as we see in the figure. For example, we have only one response for null plot 3 and 5 for Interaction effect which should lead to the extreme proportion of correct since one response would either be correct or wrong.

@@

It is also interesting to check whether the proportion correct differs if the actual plot is on the outer boundary of the lineup or inside the outer boundary. As we see in Figure \ref{fig:location_effect}, location 9, 12 are inside for Interaction effect and location 8, 12 are inside for Genotype. It does not appear to have any difference whether the actual plot is inside or outer border of the lineup.

The overall proportion of correct response does not seem to vary much for different location indicating the non-existence of location effect. Model \eqref{manova} is used to test if the mean vectors are similar for different locations. To fit the model we use anova function of stats package of \cite{R:2012}. The results are shown in Table \ref{tbl:manova}. The $p$-values for both Interaction and Genotype effect are much bigger than the conventional threshold of 0.05 and we failed to reject the hypothesis that there is no difference in location.

@@

We also fit Model \eqref{manova} with only two locations based on outer or inner plot location. By outer location we mean the data plot location which is on the border area of the lineup. For example inner locations of a lineup would be 7,8,9,12,13,14. It turns out that inner and outer locations are also not significantly different.

%% latex table generated in R 2.15.0 by xtable 1.7-0 package
%% Fri Sep  7 10:28:47 2012
%\begin{table}[hbtp]
%\caption{Fixed effect parameter estimates of generalized mixed model. Note that attempt is not significant for experiment 2. The continuous covariate, lineup difficulty, was measured by the $p$-value of the actual plot in the lineup.}
%\begin{center}
%\begin{tabular}{llrrrrl}
%  \hline
%& Parameters & Estimate & Std..Error & z.value & P-value  & \\ 
%  \hline
%\multicolumn{2}{l}{\bf{Experiment 1} } &&&& &\\
%&(Intercept) & -0.30 & 0.10 & -2.89 & 0.00 & ***\\ 
% & attempt & 0.08 & 0.02 & 4.21 & 0.00 & ***\\ 
% & lineup difficulty & -11.32 & 1.01 & -11.19 & 0.00 & ***\\ 
%\multicolumn{2}{l}{\bf{Experiment 2} } &&&&& \\
% & (Intercept) & 2.36 & 0.15 & 16.09 & 0.00 & ***\\ 
% & attempt & 0.01 & 0.03 & 0.35 & 0.72 \\ 
% & lineup difficulty & -55.03 & 2.20 & -25.00 & 0.00 & ***\\ 
%\multicolumn{2}{l}{\bf{Experiment 3} } &&&& \\
% & (Intercept) & 0.25 & 0.16 & 1.58 & 0.11 & .\\ 
% & attempt & 0.10 & 0.03 & 3.34 & 0.00 & ***\\ 
% & lineup difficulty & 3.00 & 0.56 & 5.36 & 0.00 & ***\\ 
%   \hline
%\end{tabular}
%\end{center}
%\label{tbl:model_par}
%\end{table}

% \footnote{Signif. codes: 0 Ô***Õ 0.001 Ô**Õ 0.01 Ô*Õ 0.05 Ô.Õ 0.1}

@@

This investigation allowed each observer to evaluate 10 lineups assuming that it would not cause fatigue or disinterest of the observers towards the experiment based on the pilot studies. The future research may involve doing more experiments to check if fewer or more than 10 lineups have any significant impact on performance of the observer.

@@

\section{Some figures and tables (for review)}

>>>>

Figure~\ref{fig:tower} displays a lineup of 20 plots where one of the plots is observed data, while the remaining 19 plots are rendered from data generated under a null model. Which one of the 20 plots is the most different from the others? When asked this question, 12 of 72 observers picked the same plot (with number equal to the result of $3^2+4$), with reasons given being 'asymmetry' (36\%), 'trend' (26\%) or `outliers' (16\%).  The corresponding $p$-value is 0.00023, indicating sufficient evidence to reject the null hypothesis. 
, indicating that we have enough evidence to reject the null hypothesis.
What does this mean, though? For that, we need to know the context of the data and we need to have more information about the generation of the null plots. For that, we need to know the context of the data and we need to have more information about the generation of the null plots. This example investigates the results from the 2012 US presidential election in comparison to the poll results just prior to the election. 
(Although this example is more simplistic than most of the tests conducted to date, it will serve the purpose of illustrating the lineup protocol.) The data is looking at the difference in poll results between the two (major) presidential candidates, Obama and Romney, for all states. Each panel in Figure~\ref{fig:elect-1} shows an `electoral building' \citet{nytimes} where each state in the union is represented by a rectangle. This difference is plotted horizontally, and the height of each box corresponds to the state's electoral votes. Color indicates party affiliation.  The null hypothesis is ``that the election results were consistent with the polls''. The polling results provides the null model from which data is simulated. Because each poll has a margin of error, this is used to simulate different scenarios that might have resulted on election day, if the polls were on target. A null data set is then generated as a set of draws from normal distribution, with mean equal to the difference in poll percentage of the latest state poll results, and standard deviation equal to 2.5, approximating a margin of error of 5\%. These samples are plotted as electoral buildings, and the plot made with the results from the election is placed randomly among them in a lineup of size $5\times 4$. If the null hypothesis is true the actual data plot should look just like any of the other plots, and not be identified by an observer.
 \hh{Figure~\ref{fig:tower} shows a plot of the electoral building with added context information and labels.}
 %A party needs 270 electoral votes or more to win the presidential election. 
%\begin{figure}[htbp] 
%   \centering
%   \includegraphics[width=0.8\linewidth]{electoral-5-13.pdf} 
%   \caption{Which plot looks the most different from the other plots?  Lineup \#1 - electoral building; another set of four lineups can be found in figs.~{fig:elect.2} through~{fig:elect.5}}
%   \label{fig:elect.1}
%\end{figure}

@@

\subsection{Location of Actual Plot in the Lineup} The actual data plot is placed in a random spot in a lineup. While evaluating the lineup some people may start looking from some specific part of the lineup. With the help of eye-tracking equipment \citet{zhao:2012} track the observers' eyes to see how they go through the plots in a lineup to come to their answers. The results suggest that people have particular methods of reading lineups. Some people read lineups from left to write direction while some read from upward to downward. Some people start looking from the center of the lineup while others start from the top left corner. In the earlier phase of the exercise, the observer tend to scan the plots and start comparing plots to make a final decision. Beside right-left or up-down directions observer show some diagonal movement too.

Given a specific pattern of eye movement of the observer in examining the lineup, the location of actual plot in the lineup may have some effect. Those who start exploring from the left top corner may get first glance of the actual plot if it is on that location. This should give the observer more time to compare it with the rest of the null plots. Those who start looking from the center may scan towards right or left direction and thus don't have the chance to scan the null plots continuously as they could if they would start from a corner. Thus they should scan the center plot over and over again while scanning the left or right side plots. If the data plot is in the center location it may have multiple chance to be examined and hence be identified.

@@

\paragraph{Design for Location effect:} The experiment was set up based on the gene expression data with two groups. The groups are defined based on the Genotype and Empty Vector (EV). For each of the of the Genotype, gene expression data were collected in presence or absence of iron sufficiency. Two hypothesis were of primary interest. The Genotype effect and the Genotype and iron sufficiency interaction effect.

Null plots were generated by randomizing the group structure of the data. The actual plot was randomly placed in five different locations in a lineup of size 20. The locations are 2,9,12,16,20 for Interaction effect and 1,8,12,17,20 for Genotype effect. For a lineup with specific data plot location, five different sets of null plots were used to produce 5 lineups for each location. In total we have 25 lineups for Interaction effect and 25 lineups for Genotype effect.

%We set up a Turk experiment to examine whether there is any difference of performance based on the location of actual data plot in the lineup. For this five locations of a lineup were randomly chosen to put the same actual data plot. The locations are 2,9,12,16,20 for Interaction effect and 1,8,12,17,20 for Genotype effect. For a lineup with specific data plot location, five sets of null plots are used giving 5 lineups for each location. In total we have 25 lineups for Interaction effect and 25 lineups for Genotype effect. These 50 lineups are evaluated by 100 people recruited from Amazon Mechanical Turk. Each person evaluated 3 lineups one from Interaction, one from Genotype and the third one was a test lineup. The test lineup is used to process the payment and examine the quality of the data and the responses of the test plot is not added to our analysis.

@@

Table \ref{tbl:model_result_time} presents the results of Model \eqref{eqn:trend_time}. The parameter $\alpha$ for fixed effect covariate attempt is highly significant in all the experiments. The negative estimates suggest that later attempt took less time for an evaluation. This suggest that even though observers did not improve the performance over attempt, they became efficient in responding faster. The parameter $\alpha_1$ for first attempt is also highly significant. The positive estimates of $\alpha_1$ indicates that first attempt made by an observer required much more times than other attempts. It is because for initial attempt the observer might have gone through instructions and became familiar with the experimental environment. Each page in the web site contains information about choice reason and the observers's confidence level. Also, the first page asks for observer Identification to be typed. The later pages of the web site was similar just the lineup was changed. Thus in the later attempts an observer does not need to spend any time for reading instructions. The model reflects that fact.

@@

The lineup variance of experiment 7 is estimated as $\hat{\sigma}^2 = 10.98$ from Model \eqref{eqn:trend_response}. But from Model \eqref{eqn:trend_time} it is estimated to be much smaller in all these experiments. This suggest that the harder lineups not necessarily took more time than easier lineups. The observers spent enough times to evaluate a easy lineup and for difficult lineup they might just give up at some point and gave the feedback.

To visualize how the time taken reduces over the successive attempts, we fitted Model \eqref{eqn:trend_time} excluding the covariate attempt from the model and computed the residuals. The averages of these residuals for each attempt are plotted in Figure \ref{fig:learning_trend_time} for all three experiments. Least square regression lines are fitted to the points excluding the first attempt. The downward trends are evident in the plots. All the slopes are highly significant. For attempt 1, we observed large positive residuals for each of the experiments. These results indicate that after working with the multiple tasks each observer became efficient which helped them finish the task faster in later attempts.

@@

%A plot of residuals against the attempts $A_k$ should reveal the trend if there is any. Figure \ref{fig:learning_trend} shows such plots for three experiments naming experiment 5, 6 and 7. For all the experiments we see some increasing trend of the residuals. This could be attributed to the exhaustion of subjects' attempt after some earlier evaluations are made.  But none of apparent trend are statistically significant.

%Learning trend is not only related to the proportion correct responses. It may be possible that people learn to answer fast. We examine this by the time taken for each evaluation. We fit a mixed effect linear model with time taken considering lineup and subjects as random effect. The resulting residuals are plotted in Figure \ref{fig:learning_trend_time}. It appears that in the later attempts observers take less time to evaluate a lineup. Even though it does not tell much about improving their performance over attempts, but it does indicate the improvement of their evaluation skill. Adding this information with what we observe in Figure \ref{fig:learning_trend}, we can say that the observers' improved skills could not contribute to their performance. It only help them finish the task faster in later attempts.

@@

The proportion of correct responses for each data plot location is shown in Figure \ref{fig:location_effect}. We observe some variability of performance for different sets of null plots even though same data plot was used for all these null sets. This may happen when for some set of null plots, one null plot appears to be more similar to the actual plot. In another set of null plots this may not happen making some lineups easier than others even though the actual data plot is the same. This pattern is evident in the figure as we see proportion correct for null plot 5 is consistently above the null plot 1 for each of the locations. The overall average proportions correct responses for each location are shown using dashed lines. As we see in the plot  the overall proportion is not varying much for different locations.

The sizes of the dots in  Figure \ref{fig:location_effect} represents the number of responses obtained for that location and null set. For some locations we have as many as 10 responses. For location 1, we did not have any response for null set 1 in one of the interaction lineups. The larger variability observed for Interaction lineups could be controlled at some extent by collecting more data for each location. Since the variability is larger when we have small number of responses as we see in the figure. For example, we have only one response for null plot 3 and 5 for Interaction effect which should lead to the extreme proportion of correct since one response would either be correct or wrong.

@@

We fitted Model \eqref{manova} to test if the mean performance vectors are similar for different locations. For this we use {\it anova()} function of {\it stats} package of \cite{R:2012}. The results are shown in Table \ref{tbl:manova}. The $p$-values for both Interaction and Genotype effect are much bigger than the conventional threshold of 0.05 and we failed to reject the hypothesis that there is no difference in location.

We also examined whether the proportion of correct responses differs if the actual plot is on the outer boundary or in the inner locations. The locations 7,8,9,12,13,14 in the lineup are considered inner locations and the rest of the boundary locations are considered to be outer location. As we see in Figure \ref{fig:location_effect}, location 9, 12 are inside for Interaction effect and location 8, 12 are inside for Genotype. It does not appear to have any difference whether the actual plot is inside or outer border of the lineup. We also fitted Model \eqref{manova} with two locations, inner and outer, as covariate and observed no significant differences.

@@

%
%% latex table generated in R 2.15.0 by xtable 1.7-0 package
%% Fri Sep  7 10:28:47 2012
%\begin{table}[hbtp]
%\caption{Fixed effect parameter estimates of generalized mixed model. Note that attempt is not significant for experiment 2. The continuous covariate, lineup difficulty, was measured by the $p$-value of the actual plot in the lineup.}
%\begin{center}
%\begin{tabular}{llrrrrl}
%  \hline
%& Parameters & Estimate & Std..Error & z.value & P-value  & \\ 
%  \hline
%\multicolumn{2}{l}{\bf{Experiment 1} } &&&& &\\
%&(Intercept) & -0.30 & 0.10 & -2.89 & 0.00 & ***\\ 
% & attempt & 0.08 & 0.02 & 4.21 & 0.00 & ***\\ 
% & lineup difficulty & -11.32 & 1.01 & -11.19 & 0.00 & ***\\ 
%\multicolumn{2}{l}{\bf{Experiment 2} } &&&&& \\
% & (Intercept) & 2.36 & 0.15 & 16.09 & 0.00 & ***\\ 
% & attempt & 0.01 & 0.03 & 0.35 & 0.72 \\ 
% & lineup difficulty & -55.03 & 2.20 & -25.00 & 0.00 & ***\\ 
%\multicolumn{2}{l}{\bf{Experiment 3} } &&&& \\
% & (Intercept) & 0.25 & 0.16 & 1.58 & 0.11 & .\\ 
% & attempt & 0.10 & 0.03 & 3.34 & 0.00 & ***\\ 
% & lineup difficulty & 3.00 & 0.56 & 5.36 & 0.00 & ***\\ 
%   \hline
%\end{tabular}
%\end{center}
%\label{tbl:model_par}
%\end{table}
%
% \footnote{Signif. codes: 0 Ô***Õ 0.001 Ô**Õ 0.01 Ô*Õ 0.05 Ô.Õ 0.1}

@@

This investigation allowed each observer to evaluate 10 lineups assuming that it would not cause fatigue or disinterest of the observers towards the experiment based on the pilot studies. The future research may involve doing more experiments to check if fewer or more than 10 lineups have any significant impact on performance of the observer. The learning trend in terms of time taken shows downward trends. At some points it should level of. It would be useful to know for how many lineups time taken levels off, since that would indicate the limit beyond which no learning occurs.

@@

\section{Some figures and tables (May go to Appendix)}
